---
title: "Exercises I: Statistical Modeling in R"
author: "Henrik Singmann"
date: "July 2019 (updated: `r Sys.Date()`)"
output: pdf_document
---


```{r setup, include=FALSE}
load(url("http://singmann.org/download/r/ds_vb_18.rda"))
```

# Formula Interface for Statistical Models: `~`

- Allows symbolic specification of statistical model, e.g. linear models:  
    `lm(reasoning ~ binding, ds_vb_18)`
- Everything to the left of `~` is the dependent variable.
- Independent variables are to the right of the `~`:

| Formula | &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; | Interpretation  |
| ------------------------|---|----------------------------------|
| `~ x` or `~1+x`         || Intercept and main effect of `x` | 
| ` ~ x-1` or `~0 + x`    || Only main effect of `x` and no intercept (questionable) |
| `~ x+y`                 || Main effects of `x` and `y`|
| `~ x:y`                 || Interaction between `x` and `y` (and no main effect) |
| `~ x*y` or `~ x+y+x:y`  || Main effects and interaction between `x` and `y` |


# Continuous Variables: How many Parameters in each Model?

```{r, eval=FALSE}
lm(reasoning ~ binding + updating, ds_vb_18)   # a
lm(reasoning ~ binding : updating, ds_vb_18)   # b
lm(reasoning ~ 0 + binding:updating, ds_vb_18) # c
lm(reasoning ~ binding*updating, ds_vb_18)     # d
lm(reasoning ~ 0+binding*updating, ds_vb_18)   # e
```

# Categorical Variables: How many Parameters in each Model?

```{r, eval=FALSE}
lm(reasoning ~ order, ds_vb_18)               # a
lm(reasoning ~ 0+order, ds_vb_18)             # b
lm(reasoning ~ order+training, ds_vb_18)      # c
lm(reasoning ~ 0+order+training, ds_vb_18)    # d
lm(reasoning ~ order:training, ds_vb_18)      # e
lm(reasoning ~ 0+order:training, ds_vb_18)    # f
lm(reasoning ~ order*training, ds_vb_18)      # g
lm(reasoning ~ 0+order*training, ds_vb_18)    # h
lm(reasoning ~ order+order:training, ds_vb_18)# i
```

```{r}
levels(ds_vb_18$order)    ## 2
levels(ds_vb_18$training) ## 3
```
