---
title: 'Exercise 2: Complete, No, and Partial Pooling'
author: "Henrik Singmann"
date: "January 2019 (updated: `r Sys.Date()`)"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


# Task Description

One finding reported in the literature is that the weather influences ones subjective well-being. File `weather_wb.csv` contains simulated (i.e., fictitious) data from a study in which a random sample of 60 individuals (column `id`) were asked to report their subjective well being (column `swb`) at the end of each of 45 consecutive days. Note, the start date for each participant can be a different day and participants can come from different locations/climates. Subjective well-being was a composite score constructed from four individual items. Participants had to rate how "happy", "joyful", "sad", and "angry" they felt on a scale from 0 to 10, and the last two items were reverse coded for the composite score. In addition, the data also contains the daily sun hours (column `sunh`) at the location the participant was during each day.

Your task is to analyse this data using no-pooling, at least one complete-pooling approach (ideally two different complete pooling approaches), and partial pooling using a linear mixed models. The research goal is to answer the question if there is a relationship between the daily sun hours and the reported subjective well-being. 

Do the different approaches agree in their conclusions? If not, what you think is the correct approach and why do you think the different approaches might provide a different answer? If you can, please add one figure for each of your analyses approaches to the results.

# Solution

### Load Data and Packages

Read data and load some packages.

```{r, message=FALSE}
library("tidyverse")
library("cowplot")
theme_set(theme_bw(base_size = 12) + 
            theme(legend.position="bottom", 
                  panel.grid.major.x = element_blank(),
                  panel.grid.minor.x = element_blank()))
w <- read_csv("weather_wb.csv")
```

### Descriptive Statistics

Then let's calculate some descriptives statistics. First, number of observations and participants

```{r}
w %>% 
  group_by(id) %>% 
  summarise(n = n()) %>% 
  do(psych::describe(.$n))
```

Second, descriptive statistics where data is aggregated within participants:

```{r}
w %>% 
  group_by(id) %>% 
  summarise(mean_sunh = mean(sunh),
            mean_swb = mean(swb)) %>% 
  select(-id) %>% 
  do(psych::describe(.))
```

This shows quite a considerable range in mean daily sun hours as well as mean subjective well-being.

### No Pooling


```{r}
# ...
```

### Complete Pooling 


```{r}
# ...
```

### Partial Pooling

```{r, message=FALSE}
library("lme4")
# ...
```

