---
title: "Exercise 2: Fitting Mixed Models Using lme4"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

For this exercise we will use the data of Skovgaard-Olsen et al. (2016) already used previously. We can load it directly. We also load package `lme4` for fitting some first mixed models.

```{r}
load("ssk16_dat_prepared_ex2.rda")
str(dat2)
library("lme4")
```

Variables in the data:

- `p_id`: participant identifier
- `i_id`: item identifier (i.e., id of vignette)
- `B_given_A`: original *P*(*B*|*A*)
- `B_given_A_c`: *P*(*B*|*A*) centered at midpoint of scale (as used in paper)
- `if_A_then_B`: original *P*(if *A* then *B*)
- `if_A_then_B_c`: *P*(if *A* then *B*) centered at midpoint of scale (as used in paper)
- `rel_cond`: relevance condition, irrelevant here.

Research question: Does the Equation (i.e., *P*(if *A* then *B*) = *P*(*B*|*A*)) hold? 

# Task 1: Fit 2 Models

Our general goal is to use a partial pooling approach with this data using `lm4`. We want to recreate the first two models from the lecture, add one more model, and inspect the results.

- Dependent variable: `if_A_then_B_c`
- Fixed effect: `B_given_A_c`

1. The first model only has by-participant (i.e., by `p_id`) random-slopes for `B_given_A_c`.
2. The second model has has by-participant random-slopes for `B_given_A_c` and random-intercepts, as well as the correlation between the random-effects parameters.
3. The third (and new) model only has by-participant random intercepts and no random slopes.

```{r}
# m1 <- lmer( ... , dat2)
# m2 <- lmer( ... , dat2)
# m3 <- lmer( ... , dat2)
```


Inspect the results of all models:

- Which model shows a singular fit or other convergence problems? Do you have an idea where the convergence problems come from?
- What do the parameter estimates of the random effects mean?
- Does the choice of random-effects structure affect the fixed-effect estimates and/or inferential statistics?
- Which model do you think would be the best to report?

# Task 2: Suppressing the Correlation

The only model we have not yet considered for this data is a model without correlation among the random-effects parameters. Can you set it up?

```{r}
# m4 <- lmer( ... , dat2)
```

Does this model show any benefit compared to one of the previous models?


